---
title: "Objective 2"
author: "AHM"
date: "2025-03-26"
output: html_document
---

```{r}
#reading the data set

data <- read.csv("https://raw.githubusercontent.com/CEGriffin/432-Final-Project/main/raw_data.csv")
```


```{r}
#Load the required libraries for data manipulation, modeling, and visualization.

library(dplyr)
library(tidyr)
library(tree)
library(rpart)
library(gbm)
library(randomForest)
library(ggplot2)
library(caret)
```

```{r}
#Display the first few rows, last few rows, and structure of the dataset to understand its contents.

head(data)
tail(data)
str(data)
```

```{r}
#Identify and list the column names that have missing (NA) values.

data %>% select_if(function(x) any(is.na(x))) %>% names()
```

```{r}
#Convert the population column to a factor (categorical variable).

data <- data %>%
  mutate(population = factor(population))
```

```{r}
#Get the dimensions (number of rows and columns) of the dataset.

dim(data)
```

```{r}
#Create and print a frequency table showing the distribution of different populations.

PopTable <- table(data$population)
print(PopTable)
```

splitting data
```{r}
#Remove specific non-relevant columns from the dataset to create PopData.

PopData <- data %>% 
  select(-c("Tag", "petri_dish", "family", "common_garden", "ID", "gh_bench", "gh_col", "gh_row", "Sample", "Row_Field", "Col_Field"))
```

```{r}
#list of used columns

("ChlorA", "ChlorB", "gluc_Conc", "flav_Conc", "GM_Leaf_Len", "GM_Leaf_Wid", "GM_TotalLeaf_Area", "GM_NumberOfLeaves", "Larg_Leaf_Len_Bolt", "Larg_Leaf_Wid_Bolt", "GM_StemHeight_Bolt", "GM_Leaf_Number_Bolt", "RGR1", "RGR2", "RGR3", "RGR4")
```

```{r}
# Replace the string "Dead" with "0" in all character columns of PopData.

PopData <- PopData %>%
  mutate(across(where(is.character), ~ gsub("Dead", "0", .)))

```

```{r}
# Convert the character columns (which now contain numeric values) into numeric columns.

PopData <- PopData %>%
  mutate(across(where(is.character), as.numeric))

```

```{r}
# Ensure that the population column remains a factor if it was unintentionally changed.

PopData$population <- as.factor(PopData$population)
```


```{r}
#Create a variable to split the dataset into a training set and a test set.

Test_sub <- c(1:nrow(PopData)) %%2
Train_sub <- 1-Test_sub
```

```{r}
#Split PopData into training and testing sets based on the previously defined indices.

Test <- PopData[Test_sub == 1,]
Train <- PopData[Train_sub != 1,]
```


```{r}
#Fit a decision tree model to predict population using the training data and visualize the tree.

PopTree <- tree(population ~ ., data=Train)
plot(PopTree)
text(PopTree, cex = 0.5)
```

```{r}
#Perform cross-validation and prune the decision tree to select the best tree size, then visualize it.

PopPrune <-cv.tree(PopTree, best=6, FUN = prune.tree)
plot(PopPrune)
text(PopPrune, cex = 0.5, adj=0)
```
therefore relative growth rate 1 is the most important feature for distinguishing between different populations in the data set. The RGR1 difference between populations is the most significant factor when predicting which population a particular sample belongs to. 

This suggests populations may differ in how quickly they grow relative to others. Some populations may be naturally more vigorous or have faster growth under certain conditions 


