---
title: "Objective 2"
author: "AHM"
date: "2025-03-26"
output: html_document
---

```{r}
data <- read.csv("https://raw.githubusercontent.com/CEGriffin/432-Final-Project/main/raw_data.csv")
```


```{r}
library(dplyr)
library(tidyr)
library(tree)
library(rpart)
library(gbm)
library(randomForest)
library(ggplot2)
library(caret)
```

```{r}
head(data)
tail(data)
str(data)
```

```{r}
data %>% select_if(function(x) any(is.na(x))) %>% names()
```

```{r}
data <- data %>%
  mutate(population = factor(population))
```

```{r}
dim(data)
```

```{r}
PopTable <- table(data$population)
print(PopTable)
```

splitting data
```{r}
PopData <- data %>% 
  select(-c("Tag", "petri_dish", "family", "common_garden", "ID", "gh_bench", "gh_col", "gh_row", "Sample", "Row_Field", "Col_Field"))
```

```{r}
("ChlorA", "ChlorB", "gluc_Conc", "flav_Conc", "GM_Leaf_Len", "GM_Leaf_Wid", "GM_TotalLeaf_Area", "GM_NumberOfLeaves", "Larg_Leaf_Len_Bolt", "Larg_Leaf_Wid_Bolt", "GM_StemHeight_Bolt", "GM_Leaf_Number_Bolt", "RGR1", "RGR2", "RGR3", "RGR4")
```

```{r}
# Step 1: Replace "Dead" with "0" only in character columns
PopData <- PopData %>%
  mutate(across(where(is.character), ~ gsub("Dead", "0", .)))

```

```{r}

# Step 2: Convert only relevant columns to numeric (i.e., those that are now characters with numeric data)
PopData <- PopData %>%
  mutate(across(where(is.character), as.numeric))

```

```{r}
# Step 3: Handle any remaining issues with factors or factors-like columns (like population)
# For example, if 'population' is a factor and you don't want it to be converted, you can ensure it stays a factor:
PopData$population <- as.factor(PopData$population)
```


```{r}
#assigning a training and testing set

Test_sub <- c(1:nrow(PopData)) %%3
Train_sub <- 2-Test_sub
```

```{r}
Test <- PopData[Test_sub == 0,]
Train <- PopData[Train_sub != 2,]
```


```{r}
PopTree <- tree(population ~ ., data=Train)
plot(PopTree)
text(PopTree, cex = 0.5)
```

```{r}
PopPrune <-cv.tree(PopTree, best=6, FUN = prune.tree)
plot(PopPrune)
text(PopPrune, cex = 0.5, adj=0)
```
therefore relative growth rate 1 is the most important feature for distinguishing between different populations in the data set. The RGR1 difference between populations is the most significant factor when predicting which population a particular sample belongs to. 

This suggests populations may differ in how quickly they grow relative to others. Some populations may be naturally more vigorous or have faster growth under certain conditions 


